---
output: 
    tufterhandout::html_tufte_handout
---

# 2x Numerical ~ 0

If we have two numerical variables and want to show their distribution, the obvious answer is a scatterplot.

```{r}
library(ggplot2)

df = data.frame(iq=round(rnorm(1000, 100, 15)),
                age=round(18+rgamma(1000, 2, 1)*7))
```

### Scatterplot

The scatterplot is straight-forward: one variable on x, another on y, one point for each datapoint.

```{r}
ggplot(df, aes(x=age, y=iq))+geom_point()
```

It is often useful to add some transparency to the points to make density differences apparent in high density areas.

```{r}
ggplot(df, aes(x=age, y=iq))+geom_point(alpha=0.2, size=5)
```

### 2D histogram (heat map)

We can render a 2D histogram by converting counts into colors, and binning along both x and y. 

```{r}
ggplot(df, aes(x=age, y=iq))+geom_bin2d()
```

Note that this leaves out any tiles with no observations, to avoid this, we'd need to make a new data-frame with counts for each bin, and zeros for any unobserved bins.

### 2D density plot

This is the 2D version of the density plot we saw for distributions of one numerical variable -- it smoothes data in both x and y, and then yields a contour plot.

````{r}
ggplot(df, aes(x=age, y=iq))+geom_density2d()
```

We can use the stat_density2d function directly to produce a filled density plot rather than a contour plot.

````{r}
ggplot(df, aes(x=age, y=iq))+stat_density2d(aes(fill=..level..), geom='polygon')
```

With a bit more manipulation, we can fill the whole plot area with color, if we so desire:

````{r}
ggplot(df, aes(x=age, y=iq))+stat_density2d(aes(fill=..density..), geom='tile', contour = FALSE)
```


