---
output: 
    tufterhandout::html_tufte_handout
---

# Categorical ~ Categorical

There is a subtle but important conceptual distinction between showing how the distribution of one categorical variable changes with another, explanatory, categorical variable (this page), and showing the joint distribution of two categorical variables with no explanatory variables: [2C ~ 0](2C-0.html).  Categorical~Categorical assumes that the relative proportions of the explanatory categories does not matter, and that we only care about how the response categorical variable changes with the explanatory category; these assumptions imply that the response variable is the consequence of the causal explanatory variable.  Showing the joint distribution, in contrast, conveys the relative frequencies of both variables and does not make such a causal insinuation; thus it is more appropriate if both categorical variables were observations without any experimental manipulation.

Let's start by generating the same fake bi-variate categorical data we used in [2C ~ 0](2C-0.html).

````{R}
library(ggplot2)

df.eth = data.frame(ethnicity = sample(c("white", "black", "hispanic", "asian", "native/islander"), 
                                       140, replace=T, prob=c(0.5, 0.2, 0.1, 0.15, 0.05)),
                    major = sample(c("Psychology", "Cog Sci", "Linguistics", "Business"),
                                       140, replace=T, prob=c(0.3, 0.1, 0.2, 0.4))
                    )

table(df.eth$ethnicity, df.eth$major)
````

### Adjacent stacked area bars

If we want to convey how the distribution of one categorical variable changes as a function of another, a stacked area plot if the way to go: show one categorical variable along x, and the other as a stacked area along y.  It is somewhat important in the case of C~C to normalize the different bins to show just the relative proportion, otherwise in the case of an unequal distribution of explanatory variable categories, it is tricky to see how the response variable category proportions differ.

````{r}
ggplot(df.eth, aes(x=major, fill=ethnicity))+geom_bar(position="fill", width=0.5, stat='bin')
````

Note, that here we say `position="fill"` to make the bars normalized by filling the y axis.

Note also that the implication of this plot is that major somehow influences the ethnicity distribution, which implies a particular interpretation (perhaps suggesting that majors discriminate).  You might consider it more appropriate to put ethnicity on the x axis, thus implying that ethnicity influences major selection (perhaps because you think different ethnicities choose different majors).  

````{r}
ggplot(df.eth, aes(x=ethnicity, fill=major))+geom_bar(position="fill", width=0.5, stat='bin')
````

Note also that such a relative-frequency plot obscures sample-size differences in the explanatory variable, thus making it easy to be mislead by noise due to small sample size in some categories.

Again -- these are the reasons why, when you have two observed (and not experimentally controlled) categorical variables, it is better to plot their [joint distribution](2C-0.html) without implying that one is the cause, and the other is the effect.

### Grouped histograms

Another option is to show a grouped histogram, either normalized or not.  However, I think like an unnormalized stacked area plot, this strategy makes it difficult to compare relative proporitons across explanatory categories.

````{r}
ggplot(df.eth, aes(x=ethnicity, fill=major))+geom_bar(stat='bin', position='dodge')
````