---
output: 
    tufterhandout::html_tufte_handout
---

# Facets

We discussed grids in the [ggplot introduction](ggplot.html).  Grids allow us to combine two different plots into one figure.  Facets are multiple panels of one figure which come in handy when we have too many categorical variables for us to incorporate into a single panel plot.  This is a situation that happens if we have more than two categorical variables predicting a numerical one, or if we have one or two categorical variables predicting a bivariate numerical distribution, etc.

For instance, if we want to show how a bivariate numerical relationship changes as a function of some categorical variable (2N ~ C), we could try to color-code multiple scatter-plot points in one panel, but it would generally be hard to parse.  Instead a better strategy is to arrange these 2N ~ 0 scatterplots for each category in multiple panels.

````{r}
majors = c('Psychology', 'Cog Sci', 'Business', 'Engineering', 'Homeopathy', 'Cheese Making', 'Physics')

df = data.frame()
for(major in majors ){
  r = rbeta(1,0.5,0.5)*2-1
  z = rnorm(150, 0, 1)
  w = sqrt(abs(r))
  u = r/w
  sw = sqrt(1-w^2)
  su = sqrt(1-u^2)
  a = z*w+rnorm(150,0,sw)
  b = z*u+rnorm(150,0,su)
  df = rbind(df, data.frame(IQ=round(a*15+100), GPA=pmax(0,pmin(4,b*2+2.8)), major=major))
}
str(df)
````

If we try to plot the IQ-GPA relationship for each of the five majors in a single panel, color-coding by major, we get a mess:

````{r}
library(ggplot2)
ggplot(df, aes(x=IQ, y=GPA, color=major))+geom_point()
````

However, if we arrange these in different panels ("facets"), it is much easier to see what is going on.

````{r}
ggplot(df, aes(x=IQ, y=GPA))+geom_point()+facet_wrap(~ major)
````

### facet_wrap

`facet_wrap` is useful when we have one categorical variable that has many different categories, thus we want the panels for that one variable to 'wrap' (spanning multiple rows and columns).  We can alter how many rows and columns we have by setting either `nrow` or `ncol` in facet wrap.

````{r}
ggplot(df, aes(x=IQ, y=GPA))+geom_point()+facet_wrap(~ major, ncol=4)
````

### facet_grid

`facet_grid` is useful when we want to partition panels along two orthogonal categorical variables, thus making one variable correspond to rows and another to columns.

````{r}
majors = c('Psychology', 'Cog Sci', 'Business', 'Engineering', 'Homeopathy', 'Cheese Making', 'Physics')
sexes = c('male', 'female')

df = data.frame()
for(major in majors ){
  for(sex in sexes){
    r = rbeta(1,0.5,0.5)*2-1
    z = rnorm(150, 0, 1)
    w = sqrt(abs(r))
    u = r/w
    sw = sqrt(1-w^2)
    su = sqrt(1-u^2)
    a = z*w+rnorm(150,0,sw)
    b = z*u+rnorm(150,0,su)
    df = rbind(df, data.frame(IQ=round(a*15+100), GPA=pmax(0,pmin(4,b*2+2.8)), major=major, sex=sex))
  }
}
str(df)
````

For instance, here we want to look at the IQ-GPA relationship for all the combinations of sex and major, arranging sex along the rows, and major along the columns:


````{r fig.width=12, fig.height=6}
ggplot(df, aes(x=IQ, y=GPA))+geom_point()+facet_grid(sex ~ major)
````


