---
  output: 
    html_document
---
  
# Anscombe's quartet

The data for [Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet) are built into R, and it is a somewhat useful exercise to reshape the data frame and produce the plot.

```{r}
str(anscombe)
```

Note that these data are in "wide" format, we have 4 x columns and 4 y columns, rather than having an x column, a y column, and a column indicating whether the values correspond to set 1, 2, 3, or 4.  We will reshape these data.

```{r}
# let's use tidyr and dplyr to reshape this into long format.
library(tidyr)
library(dplyr)
library(ggplot2)
# first, we add a new variable (observation number)
anscombe.tidy = mutate(anscombe, observation=seq_len(n()))
# now we reshape into (too) long format (all x and y values in one column)
anscombe.tidy = gather(anscombe.tidy, key, value, -observation)
# now we separate the key variable ("x1", "x2", .. "y1") into two variables:
# "variable" ("x" or "y") and "set" (1, 2, 3, 4)
anscombe.tidy = separate(anscombe.tidy, 
                         key, 
                         c("variable", "set"), 
                         1, 
                         convert = TRUE)
# Now we change the "set" to be a character variable
anscombe.tidy = mutate(anscombe.tidy, 
                       set = c("I", "II", "III", "IV")[set])
# Finally, we spread into a wide format x and y values:
anscombe.tidy = spread(anscombe.tidy, variable, value)

str(anscombe.tidy)
```

Great, now these data are in "long" format, and we can easily handle them in ggplot to produce a multi-panel plot.

```{r, fig.height=3.5}
ggplot(anscombe.tidy, aes(x=x,y=y))+facet_grid(.~set)+
  geom_smooth(method='lm', formula=y~x)+
  geom_point(color="black", fill="orange", shape=21, size=3.5)+theme_minimal()+theme(text=element_text(size=16))
```

What's notable about Anscombe's quartet is that their summary statistics are so similar.

```{r}

anscombe.tidy %>% 
  group_by(set) %>%
  summarize(
      n=n(),
      mean.x=round(mean(x),2), 
      mean.y=round(mean(y),2), 
      sd.x=round(sd(x),2), 
      sd.y=round(sd(y),2), 
      correlation=round(cor(x,y),2))
```

These matching ns, means, standard deviations, and correlation, means that the slope and intercept, and corresponding statistical tests, are all equivalent; yet the data look clearly different.  

The point here is: **summary statistics of marginal distributions and linear relationships necessarily overlook some (potentially important) aspects of the data.**
