---
output: 
    html_document
---

```{r Setup, include=FALSE, results='hide', warning=FALSE}
library(knitr)
opts_chunk$set(dev=c("png","pdf"),
               fig.width=10,
               fig.height=5,
               dpi=300,
               fig.show="hold",
               fig.lp="fig:",
               cache=TRUE,
               par=TRUE,
               echo=TRUE,
               message=FALSE,
               warning=FALSE)
library(ggplot2)
my_theme <- theme_bw() + 
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "transparent", color=NA),
        plot.background = element_rect(fill="transparent", color=NA),
        axis.text = element_text(size=14),
        title = element_text(size=18),
        axis.title = element_text(size=16))
```

# (Student's) t-tests

We use t-tests when we calculate the standard error based on the sample standard deviation.  The logic of why we must use the t-distribution in these cases is described when we consider the [sampling distribution of the sample variance, and the t statistic](nhst-t-distribution.html).  Read that first, and then we can quickly cover the slightly different t-test varieties we can use to compare sample means.

## 1-sample t-test

So if we have a sample, which we assume to be normally distributed with unknown population mean and variance, and want to test whether that sample has a mean different from some null hypothesis mean ($\mu_0$).  We would proceed as follows: estimate the sample mean and sample variance, and then compute a t-statistic:

$t_x = (\bar x - \mu_0) / (s_x / \sqrt n)$

Under the null hypothesis this t statistic will follow a t distribution with one parameter, the degrees of freedom: $df = n-1$.  

The degrees of freedom parameter changes the t-distribution depending on our certainty in our estimated sample variance.  If $n$ is very large, then we should be very certain that our estimated sample variance will not be very different from the population variance.  Consequently, the t-statistic when $n$ is large will be distributed no differently than a z-statistic: it will follow a normal distribution.  However, when $n$ is small, we have substantial uncertainty about the estimated sample variance, so the distribution of the t-statistic will take on a different shape.  Specifically, it will have heavier tails (it will have greater kurtosis), this should be intuitive because if we underestimate the population standard deviation, our t-statistic (which has estimated standard deviation in the denominator) might be quite large by chance -- hence the heavier tails.

Anyway, so to compute a p-value for a t-test, you would look up your t-statistic in the cumulative t-distribution (just as p-values for a z-test are obtained from a cumulative standard normal distribution), with $n-1$ degrees of freedom:

For a conventional two-tailed test, we would calculate the p-value as:

`2*pt(-abs(tx),n-1)` (notice that we are merely substituting `pt` for `pnorm`)

If the degrees of freedom are small, the p value from the t distribution may be considerably larger than for a normal distribution:

`2*pt(-abs(3),2)`=`r 2*pt(-abs(3),2)`   
compare to   
`2*pnorm(-abs(3))`=`r 2*pnorm(-abs(3))`   

However, in `R` we can do this quickly using the `t.test` command.

```{r}
df = data.frame(our.iqs = rnorm(15, 103,15))
(test.results = t.test(df$our.iqs, mu = 100))
```

We can access the various aspects of the test result as:

```{r}
test.results$statistic # the t statistic
test.results$parameter # the degrees of freedom
test.results$p.value   # the p-value
```

## Paired / repeated-measures t-test

In other cases we might have two measurements from a single source.  The canonical example of this would be "before" and "after" measurements.  e.g., for each person we measure their weight before they go on our new-fangled diet, and after 3 months of being on the diet, so for each person we have two measurements: before and after the 3 months of dieting.  We might then want to see whether people lost weight, or gained weight, or didn't change between the two tests.

This scenario offers us an opportunity to reduce person-to-person variability by calculating the difference of before and after scores, and then calculating a 1-sample t-test on the difference, comparing them to a mean of 0:

```{r}
x = rnorm(10, 170, 10)
df = data.frame(before = x+rnorm(10, 5, 2),
                after = x+rnorm(10, 0, 2))
df$difference = df$after - df$before
(test.results = t.test(df$difference, mu=0))
```

By calculating the before-after difference, we can effectively get rid of the across-subject variation, which will be manugest in both the before and after weights.  This is the virtue of repeated-measures designs in general.  For t-tests, such a design only makes sense if we have a "paired" design (such that our data come in pairs; here: before/after).

## 2-sample, presumed equal variance, t-test

Another setting is that we have two independent samples ($x$ and $y$) and want to know if they differ in their means, under the assumption that they have the same standard deviation.  This kind of setting may arise if we have a sample of males and a sample of females, and want to see if their heights are systematically different:

In R this can be run quickly as:

```{r}
df = rbind(data.frame(height = rnorm(10, 70, 5), sex="male"),
           data.frame(height = rnorm(18, 65, 5), sex="female"))
(test.results = t.test(df$height[df$sex=="male"], 
                       df$height[df$sex=="female"], 
                       var.equal = T))
```

Note that:

(a) we must specify that we assume the variances are equal (the `var.equal=T` portion)    
(b) the degrees of freedom are equal to $n_x + n_y -2$.   
(c) the standard error is calculated by pooling variance around the mean of x and the mean of y.   


## 2-sample, unequal variance, t-test

The final model we might have of our 2 samples of data is that they come from normal distributions with different means and (*potentially*) different variances, and we are interested in testing whether their means are equal.

If we do not assume equal variances, we can run the "Welch's t-test".  This is the default behavior of R's t.test function:


```{r}
df = rbind(data.frame(val = rnorm(10, 105, 20), variable="x"),
           data.frame(val = rnorm(18, 98, 10), variable="y"))
(test.results = t.test(df$val[df$variable=="x"], df$val[df$variable=="y"]))
```

Note that:

(a) we must specify that the default is to *not* assume the variances are equal (default is `var.equal=F`; this is the deault behavior in R)    
(b) the degrees of freedom is some weird non-integer value         

It is worth getting an intuition for what the degrees of freedom are doing in a Welch's t-test:     
The unequal variance degrees of freedom are somewhere between $n - 1$ and $n_x + n_y -2$, where $n$ is the size of the sample with the larger standard error of the mean).  What this does is return roughly the two-sample equal-variance degrees of freedom ($df =  n_x + n_y -2$) when the standard errors are approximately equal (since in that case you are essentially doing an equal-variance t-test), and returns the one-sample degrees of freedom ($n-1$), when one variance is much larger than the other.  The latter should make sense: one sample is much more variable than another, there is no sampling variability in the precise sample, so we are effectively comparing the noisy sample to a determinate value -- a one-sample t-test.

## Power calculations.

Instead of doing algebra to calculate power for a t-test, as we did for a normal, we will just use the `pwr::` library.

```{r}
library(pwr)
# get power for a particular sample size...

# for a 1 sample (or repeated measures) t-test
pwr::pwr.t.test(n = 20, d=0.5, sig.level=0.05, type = 'one')

# for a 2 sample t-test (equal sample sizes, each group has sample size n)
pwr::pwr.t.test(d=0.5, n=10, sig.level=0.05, type='two')

# for a 2 sample t-test (unequal sample sizes)
pwr::pwr.t2n.test(d=0.5, n1=10, n2=10, sig.level=0.05)

# get sample size for a particular level of power

# fir a 1 sample (or repeated measures) t-test
pwr::pwr.t.test(d = 0.5, power = 0.8, sig.level=0.05, type = 'one')

# for a 2 sample t-test (find one sample size for both groups)
pwr::pwr.t.test(d=0.5, power=0.8, sig.level=0.05, type='two')
```


## Summary of tests for the mean and effect sizes

| Test  |	test statistic | d.f. | effect size |   
|:------|:----:|:----:|:----:|          
| Z-test |  $z = \frac{\bar x - \mu_0}{\sigma_0 / \sqrt n}$ | na | $\hat d = \frac{\bar x - \mu_0}{\sigma_0}$ |   
| t-test | $t = \frac{\bar x - \mu_0}{s_x / \sqrt n}$ | $n-1$ | $\hat d = \frac{\bar x - \mu_0}{s_x}$ |    
| paired t-test | $t = \frac{\bar d - \mu_0}{s_d / \sqrt n}$ | n-1 | $\hat d = \frac{\bar d - \mu_0}{s_d}$ |    
| eq. var t-test | $t = \frac{\bar y - \bar x}{s_p \sqrt{1/n_x + 1/n_y}}$ | $n_x+n_y-2$ | $\hat d = \frac{\bar y - \bar x}{s_p}$ |    
| uneq. var t-test | $t = \frac{\bar y - \bar x}{\sqrt{s_x^2/n_x + s_y^2/n_y}}$ | \*\* | $\hat d = \frac{\bar y - \bar x}{\sqrt{s_x^2 + s_y^2}}$ |

\*\* The degrees of freedom for an unequal variance t-test are given by the following long expression (which you should not try to learn):    
$df = \frac{(s_x^2/n_x + s_y^2 / n_y)^2}{sx^4/(n_x^2(n_x-1)) + s_y^4/(n_y^2(n_y-1))}$

## Math.

This section is here for thoroughness, but there is no particularly good reason to commit this stuff to memory.

### Math behind 2-sample equal variance t-test

Specifically, our model is:

$x \sim \text{Normal}(\mu_x, \sigma)$     
$y \sim \text{Normal}(\mu_y, \sigma)$

Note that here we have assumed that x and y have the same standard deviation.  For reasons that become clear later, this assumption is usually made when running a t-test, especially when calculating one by hand.

Under this assumption, our expected distributions for the sample means will be:   
$\bar x \sim \text{Normal}(\mu_x, \sigma / (\sqrt n_x))$    
$\bar y \sim \text{Normal}(\mu_y, \sigma / (\sqrt n_y))$

And the distribution of the difference between the means ($\bar y - \bar x$), will be:    
$(\bar y - \bar x) \sim \text{Normal}(\mu_y - \mu_x, \sqrt{\frac{\sigma^2}{n_x} + \frac{\sigma^2}{n_y}})$     
(the random variable corresponding to the difference of $\bar x$ and $\bar y$ will have a variance equal to the sum of the variances $\sigma^2_{\bar x} = \sigma^2/n_x$ and $\sigma^2_{\bar y} = \sigma^2/n_x$)

The standard deviation of the sampling distribution of the difference of the means (this is the "standard error" of the difference between two means), $\sqrt{\frac{\sigma^2}{n_x} + \frac{\sigma^2}{n_y}}$, can be simplified to: $\sigma \sqrt{1/n_x + 1/n_y}$.

In this case, as usual, we want to test a null hypothesis, and here the null hypothesis is specified as the difference between the two means $H_0 = (\mu_{y_0} - \mu_{x_0})$.  This is usually 0: our null hypothesis is usually that the two means do not differ.  

Since we do not know the shared standard deviation, $\sigma$, we have to estimate it.  To do so, we have to "pool" information from both $x$ and $y$, since they should both inform our estimate of the shared $\sigma$.

We can compute $\bar x$ and $\bar y$, as well as $s_x$ and $s_y$, however, our estimate of the common $\sigma$ must pool $s_x$ and $s_y$.  This pooled estimate is usually written as $s_p$, and is a *weighted average* of sample variances:

$s_p^2 = \frac{\sum_i (n_i - 1) s_i^2}{\sum_i (n_i -1)}$

Since we only have two groups here, this general statement about how to pool variances boils down to:

$s_p^2 = \frac{(n_x-1) s^2_x + (n_y-1) s^2_y}{n_x + n_y - 2}$

(Note that these equations refer to pooled *variance*, so the pooled standard deviation would be the square root of these!)

Since this is our estimated variance, we can use the standard error formula we derived before to compute the estimated standard error of the difference between two means:

$s_{\bar y - \bar x} = s_p \sqrt{1/n_x + 1/n_y}$.

Once we have obtained a standard error, we can compute a t-statistic.

$t_{\bar y - \bar x} = \frac{(\bar y - \bar x) - (\mu_{y0} - \mu_{x0})}{s_{\bar y - \bar x}}$

Sometimes, for clarity, this is written out as the equivalent expression:

$t_{\bar y - \bar x} = \frac{(\bar y - \bar x) - (\mu_{y0} - \mu_{x0})}{s_p \sqrt{1/n_x + 1/n_y}}$

Usually, the null hypothesis $(\mu_{y0} - \mu_{x0})$ is 0: meaning that the difference between the means ought to be 0, so the t-statistic ends up being shortened to:

$t_{\bar y - \bar x} = \frac{(\bar y - \bar x)}{s_p \sqrt{1/n_x + 1/n_y}}$

This t-statistic will follow the t-distribution with degrees of freedom matching the denominator in the pooled sample variance calculation:   

$df = n_x + n_y -2$ 

#### Special case of equal sample sizes

Sometimes textbooks present special formulas for cases where $n=n_x=n_y$ (where the sample sizes are equal), but those formulas can be calculated directly from the more general formulas that assume unequal sample sizes by plugging $n$ in for both $n_x$ and $n_y$:     
$df = 2n-2$   
$s_p^2 = (s_x^2 + s_y^2)/2$   
$s_{\bar y - \bar x} = s_p \sqrt{2/n}$

### Math behind unequal variance t-test

Specifically, the model we consider is:
  
$x \sim \text{Normal}(\mu_x, \sigma_x)$   
$y \sim \text{Normal}(\mu_y, \sigma_y)$

Under this assumption, our expected distributions for the sample means will be:   
$\bar x \sim \text{Normal}(\mu_x, \sigma_x / (\sqrt n_x))$    
$\bar y \sim \text{Normal}(\mu_y, \sigma_y / (\sqrt n_y))$

And the distribution of the difference between the means ($\bar y - \bar x$), will be:    
$(\bar y - \bar x) \sim \text{Normal}(\mu_y - \mu_x, \sqrt{\frac{\sigma_x^2}{n_x} + \frac{\sigma_y^2}{n_y}})$

So we can proceed to calculate $\bar x$, $\bar y$, $s_x$, $s_y$, and plug these in to get a t-statistic:

$t_{\bar y - \bar x} = \frac{(\bar y - \bar x)}{\sqrt{s_x^2/n_x + s_y^2/n_y}}$.

(Again we have dropped the $ - ($\mu_{y0} - \mu_{x0}$)$ from the numerator because in most cases in which this test is applied $(\mu_{y0} - \mu_{x0})=0$.)

However, the trickiness comes in defining the degrees of freedom for this t-test.  The trickiness arises from the fact that we have two estimated standard deviations; how do we represent the sampling variability associated with both?

This was a hairy problem when folks were first dealing with t-tests, until the reasonably accurate approximation via ``Welch's t-test", which defines the degrees of freedom for this test statistic as:

$df = \frac{(s_x^2/n_x + s_y^2 / n_y)^2}{sx^4/(n_x^2(n_x-1)) + s_y^4/(n_y^2(n_y-1))}$

This unwieldy definition of the degrees of freedom generates all the desirable properties for the t-test.  Intuitively, what this formula does is return    
$n -1 \leq df \leq n_x + n_y -2$    
(where $n$ is the size of the sample with the larger sample variance).  Intuitively, what this does is return roughly $df =  n_x + n_y -2$ when $s_x$ and $s_y$ are approximately equal (since in that case you are essentially doing an equal-variance t-test), and this formula returns $n_x-1$ when $s_x$ is much larger than $s_y$ (and vice versa; this should make sense: when $s_x$ is really big compared to $s_y$, this t-test is equivalent to doing a 1-sample t-test of $x$ vs the point-null hypothesis of $\mu_0 = \mu_y$).
