---
output: html_document
---

# Visualizations

These notes are largely there to provide quick examples of ggplot2 code that generates graphs for particular data.  I assume that the ggplot2 introduction from R4DS got you oriented, but if not, we have a few more notes on the basics of [ggplot](vis.ggplot.html).

## General rules for scientific data visualization.

The order of these rules indicates their priority (as I see it): rules further down are superceded by higher up rules if they are in conflict.

1. Everything should be labeled, and interpretable without consulting a figure caption or having to solve a puzzle.

1. Graphs should facilitate *relevant* quantitative interpretation and comparisons.

1. Graphs should represent variability and uncertainty to permit inferential statistics by eye

1. Graphs should follow conventions for the kind of information/data being presented.

1. Graphs should not waste ink and should otherwise look pretty.

## Picking a plot (what's convention)

When you make a plot, you are trying to show the relationship between one or more response/outcome variables, and some explanatory variables (in experimental settings, these are often called dependent and independent variables, respectively).  These variables can be classified as either categorical or numerical.  I find it helpful to think of conventional plot options in terms of a pseudo formula: what kind of response variable is being explained by what types of explanatory variables.  e.g., categorical ~ numerical would mean you are showing how a categorical response variable changes with some numerical explanatory variable.  Most of this document is structured based on such formulas.

## Categorical ~ 0

How do we show the distribution of some categorical variable (that's what I mean by "~0": no explanatory variables)?  The three main plot types to consider here are: histogram, pie chart, and stacked area plot.

Histograms make comparisons of frequency across categories easiest (you just have to compare how tall two bars are).  Pie charts make this somewhat difficult, because we are pretty bad at visually comparing similar angles; however, pie charts do effectively convey the absolute proportion of a category (which is hard to assess in a histogram).  Stacked area plots are a sort of compromise of both: comparisons across categories are harder than in a histogram, but easier than a pie chart; absolute proportion is easier than in a histogram, but harder than a pie chart.


For our plots of a categorical variable with no expanatory variables, we will use the SPSP demographics data.

````{r, warning=F, message=F}
library(tidyverse)
spsp <- read_csv('http://vulstats.ucsd.edu/data/spsp.demographics.cleaned.csv')
spsp %>% group_by(ethnicity) %>% summarize(n=n()) %>% knitr::kable()
````

### Histogram

The classic distribution plot is a histogram: one bar per category, usually arranged along the x-axis, with the y-axis showing frequencies.

````{r, warning=F, message=F, fig.width=3, fig.height=4}
( plot1 <- ggplot(spsp, aes(x=ethnicity, fill=ethnicity)) + 
  geom_bar()+
  ## the lines above form the basis of the plot, 
  ## the lines below just make it look nicer
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90, hjust=1)) )
````

### Pie chart

Pie charts show the proportion of each category as wedges on a circle.  (Pie charts are generally [frowned upon](http://www.businessinsider.com/pie-charts-are-the-worst-2013-6), so you better have a really good reason to use them.)  Also, because they are out of favor, making them in ggplot is a bit weird: you have to make a stacked area plot, then convert the coordinates to polar.

````{r, warning=F, message=F, fig.width=1.5, fig.height=1.5}
( plot2 <- 
  ggplot(spsp, aes(x=factor(1),fill=ethnicity))+
  geom_bar(aes(y = (..count..)/sum(..count..)), width=1)+
  coord_polar(theta="y")+
  ## the lines above form the basis of the plot, 
  ## the lines below just make it look nicer
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(legend.position = "none",
        axis.title = element_blank(),
        axis.text = element_blank(),
        panel.grid = element_blank()) )
````


### Stacked area

Think of a stacked area plot as a histogram, with all the bars stacked on top of each other (hopefully, with different colors!).  This plot connects to methods of plotting a categorical variable as a function of some other variables.

````{r, warning=F, message=F, fig.width=4, fig.height=2}
( plot3 <- ggplot(spsp, aes(x=factor(1),fill=ethnicity))+
  geom_bar(width=0.75, stat='count')+
  ## the lines above form the basis of the plot, 
  ## the lines below just make it look nicer
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  guides(fill=guide_legend(title="SPSP\nethnicity"))+
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank()) )
````

## numerical ~ 0

To look at the distribution of a numeric variable, we will generally need to do something to smooth or bin over the number line*.  Histograms have explicit bins, and show the frequency of numbers in each bin.  Density plots smooth over the numberline not in explicit bins, but with a continuous kernel.  Histograms do a better job of coneying how noisy your data are, but are quite sensitive to bin placement: they will give you a different impression depending on bin width and bin shifts (e.g., bin 1-2, 2-3, ... vs 0.5-1.5, 1.5-2.5, ....).  Because density plots do not have discrete bins, they are smoother, so there is no notion of shifting a bin, and they tend to be less sensitive to bandwidth; however, they tend to make your data look clean even when they are not.

Let's use some data from a 10 mile race to plot the distribution of a numeric variable.

```{r, warning=F, message=F}
load(url("http://vulstats.ucsd.edu/data/cal1020.cleaned.Rdata"))
glimpse(cal1020)
```

### Histogram & density

```{r, warning=F, message=F, fig.width=6, fig.height=2.5}
plot1 <- ggplot(cal1020, aes(x=time.sec/60))+
  geom_histogram(binwidth = 1)+
    ggtitle('histogram')+
  scale_x_continuous('Minutes', breaks=seq(0,240, by=30))+
  theme_minimal()

plot2 <- ggplot(cal1020, aes(x=time.sec/60))+
  geom_density(fill='gray', alpha=0.5)+
    ggtitle('density')+
  scale_x_continuous('Minutes', breaks=seq(0,240, by=30))+
  theme_minimal()
gridExtra::grid.arrange(plot1, plot2, nrow=1)
```

## numerical ~ categorical

This is the most common data visualization category I see in psychology: how does some numerical variable change as a function of condition (category).  Usually it is portrayed as a bar plot (hopefully with error bars).  But we can do better.

First, let's make all the variations, and put them in one figure (below we just make each plot, and then show them all together.)

### Bar plot with error bars

We adopt the common approach of using the mean +/- 1 standard error of the mean as the error bars here.

```{r, warning=F, message=F}
plot1 <- ggplot(cal1020, aes(x=sex, fill=sex, y=speed.mph))+
  stat_summary(fun.y = mean, 
               geom="bar")+
  stat_summary(fun.data = mean_se,
               geom="errorbar", 
               width=0.5)+
  scale_y_continuous('Avg speed (mph; +/- s.e.m.)', breaks = seq(0, 15, by=1))+
  ggtitle('Barplot')+
  theme_minimal()+
  coord_cartesian(ylim=c(5,7))+
  theme(legend.position = 'none')
```

### Jittered data points.

Data points are jittered, because otherwise they would fall on top of each other, and would be impossible to discern.

```{r, warning=F, message=F}
plot2 <- ggplot(cal1020, aes(x=sex, color=sex, y=speed.mph))+
  geom_jitter(size=0.1, alpha=0.5)+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  ggtitle('Jittered')+
  theme_minimal()+
  theme(legend.position = 'none')
```

### Viola/Violin plot

Like a smoothed density histogram, except it's symmetric, so it looks nicer when many are arrayed side by side.

```{r, warning=F, message=F}
plot3 <- ggplot(cal1020, aes(x=sex, fill=sex, y=speed.mph))+
  geom_violin()+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  ggtitle('Violin')+
  theme_minimal()+
  theme(legend.position = 'none')
```

### Box and whiskers plot

This is a visualization of a bunch of summary statistics of the distribution.  By default, these summary statistics are: the median (middle line), the 25th and 75th percentile (edges of the box), 25th percentile - 1.5(IQR), and 75th percentile + 1.5(IQR) (the whiskers); and it shows the "outliers" (data points that are beyond those IQR intervals.

```{r, warning=F, message=F}
plot4 <- ggplot(cal1020, aes(x=sex, fill=sex, color=sex, y=speed.mph))+
  geom_boxplot(alpha=0.5, outlier.alpha = 0.1)+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  ggtitle('Boxplot')+
  theme_minimal()+
  theme(legend.position = 'none')
```

###  Overlayed densities

(here we flip the coordinates so we can picture them along side the other graphs)

```{r, warning=F, message=F}
plot5 <- ggplot(cal1020, aes(x=speed.mph, fill=sex, color=sex))+
  geom_density(alpha=0.5)+
  coord_flip()+
  scale_x_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  ggtitle('Densities')+
  theme_minimal()+
  theme(legend.position = 'none')
```

###  Empirical cumulative distribution

(here we flip the coordinates so we can picture them along side the other graphs)

```{r, warning=F, message=F}
plot6 <- ggplot(cal1020, aes(x=speed.mph, fill=sex, color=sex))+
  stat_ecdf(geom='line', size=1, alpha=0.75)+
  coord_flip()+
  scale_x_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  scale_y_continuous('CDF', breaks=c(0, 0.5, 1.0))+
  ggtitle('ECDF')+
  theme_minimal()+
  theme(legend.position = 'none')
```

### Comparisons

```{r, warning=F, message=F, fig.width=8, fig.height=3}
gridExtra::grid.arrange(plot1, plot2, plot3, plot4, plot5, plot6, nrow=1)
```

The mean + standard error bar plot makes the implicit statistical comparison (t-test) easy to do by eye, but it obscures what the actual data look like (both the underlying variability, as well as it's messiness).  The jittered plot is very faithful to the underlying data, but it's tricky to figure out how the distributions compare.  The violin plot hides some of the data messiness, but does make comparisons easier.  The boxplot is useful for showing the five-number summary, but not much else.  Overlayed density plots are more intuitive than violin plots, but they only work well for a small number of categories.  Empirical CDFs are also only good for a few categories, and their strength is in showing differences in the tails of the distributions.

### Recommendations 

My suggestion is to use a point+range to indicate the mean and standard error (thus facilitating comparisons), and to overlay that on some representaiton of the underlying data.  I recommend using a jittered display of individual data points if you have relatively little data, and a violin plot if you have a lot of data (such that the jitter is very hard to make informative).

Below, to show the few-data-points case, I sub-sample the data.

```{r, warning=F, message=F, fig.width=5, fig.height=3}
# recommendation for few data points
plot1 <- cal1020 %>% 
  sample_n(50) %>%
  ggplot(aes(x=sex, fill=sex, color=sex, y=speed.mph))+
  geom_jitter(width=0.25, size = 0.75, alpha=0.7)+
  stat_summary(fun.data = mean_se,
               geom="pointrange", 
               fatten = 2,
               size=1)+
  scale_y_continuous('Speed (mph +/1 sem)', breaks = seq(0, 15, by=1))+
  theme_minimal()+
  theme(legend.position = 'none')

## Here we will write a function to generate a 95% confidence interval
mean_ci <- function(x){
  m = mean(x)
  se = sd(x)/sqrt(length(x))
  ql = qnorm(1-0.025)
  c('y'=m, 'ymin'=m-ql*se, 'ymax'=m+ql*se)
}
plot2 <- cal1020 %>% 
  ggplot(aes(x=sex, fill=sex, color=sex, y=speed.mph))+
  geom_violin(alpha=0.3)+
  stat_summary(fun.data = mean_ci,
               geom="pointrange",
               fatten = 2,
               size=1)+
  scale_y_continuous('Speed (mph +/- 2 sem)', breaks = seq(0, 15, by=1))+
  theme_minimal()+
  theme(legend.position = 'none')

gridExtra::grid.arrange(plot1, plot2, nrow=1)
```


## numerical ~ numerical (2 x numerical ~ 0)

To show the joint distribution of two numerical variables, or one numerical variable as a function of another, the obvious choice is a scatterplot.  However, a scatterplot, just like a jittered display of data points, is difficult to make usable when you have a lot of data (you need to tinker with point size and point transparency so that the display isn't just an undifferentiated blob).  A 2D histogram, mapping frequency onto color, is a better choice if you have a lot of data.

### Scatter and heatmap

```{r, warning=F, message=F, fig.width=6, fig.height=3.5}
plot1 <- cal1020 %>%
  ggplot(aes(x=age, y=speed.mph, alpha=0.5)) + 
  geom_point(size=0.5, position=position_jitter(0.25))+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  ggtitle('scatterplot')+
  theme_minimal()+
  theme(legend.position = 'none')

plot2 <- cal1020 %>%
  ggplot(aes(x=age, y=speed.mph)) + 
  geom_bin2d()+
  scale_fill_continuous(low="#DDDDDD", high="#000000")+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  ggtitle('heatmap histogram')+
  theme_minimal()
  
gridExtra::grid.arrange(plot1, plot2, nrow=1)
```

### Conditional means

The other popular way to show the relationship between a numerical response (y) and a numerical explanatory variable (x) is to show the mean (plus error bars) of y for a given range of x.  Such conditional means are typically connected with a line, and sometimes even the error bars are connected in a ribbon.  This is the most direct way of showing the "conditional" means, but is generally impractical unless you have a lot of data.  The most common way to show conditional means is to show the fitted y~x line; this takes us further away from the data (because it assumes that the conditional means follow a line), and usually gives us a false impression of how linear the relationship is.  The other common alternative is to do "LOESS" smoothing (locally weighted regression); for each point we are plotting, the LOESS method fits a function to all the data, but weights the data based on how far away it is from the point we are interested; the consequence is a smooth, wiggly line that generally looks nice, but often misleads folks into thinking that they have evidence for non-linearities in their data.

Generally, I would recommend showing the raw data distribution (as in the subsection above), plus a representation of whatever model you fit to those data (if you fit a line, show a line; if you fit a parabola, show the parabola).

```{r, warning=F, message=F, fig.width=8, fig.height=3}
## make bins and calculate mean, se
cal1020.binned <- cal1020 %>% 
  group_by(age.bin = floor(age/5)*5+2.5) %>%
  summarize(mean = mean(speed.mph), sem=sd(speed.mph)/sqrt(n()))
  
plot1 <- cal1020.binned %>% 
  ggplot(aes(x=age.bin, y=mean, alpha=0.5)) + 
  geom_point()+
  geom_line()+
  geom_pointrange(aes(y=mean, ymin=mean-sem, ymax=mean+sem))+
  scale_y_continuous('Mean speed (mph +/- sem)', breaks = seq(0, 15, by=1))+
  ggtitle('y|x')+
  theme_minimal()+
  theme(legend.position = 'none')

plot2 <- cal1020.binned %>% 
  ggplot(aes(x=age.bin, y=mean, alpha=0.5)) + 
  geom_line()+
  geom_point()+
  geom_ribbon(aes(ymin=mean-sem, ymax=mean+sem))+
  scale_y_continuous('Mean speed (mph +/- sem)', breaks = seq(0, 15, by=1))+
  ggtitle('y|x Ribbon')+
  theme_minimal()+
  theme(legend.position = 'none')

plot3 <- cal1020 %>%
  ggplot(aes(x=age, y=speed.mph, alpha=0.5)) + 
  geom_smooth(method='lm', color='red', fill='red', alpha=0.5)+
  scale_y_continuous('Fitted speed (mph + 95% CI)', breaks = seq(0, 15, by=1))+
  ggtitle('Line')+
  theme_minimal()+
  theme(legend.position = 'none')

plot4 <- cal1020 %>%
  ggplot(aes(x=age, y=speed.mph, alpha=0.5)) + 
  geom_smooth(method='loess', color='blue', fill='blue', alpha=0.5)+
  scale_y_continuous('Fitted speed (mph + 95% CI)', breaks = seq(0, 15, by=1))+
  ggtitle('LOESS')+
  theme_minimal()+
  theme(legend.position = 'none')

gridExtra::grid.arrange(plot1, plot2, plot3, plot4, nrow=1)
```

### numerical ~ numerical + categorical

If you add a categorical variable to those numerical~numerical plots, the obvious choice would be to color the points/fitted lines.  However, if you have a lot of data, the scatterplots become unusable, so sometimes it is worth splitting the different categories into facets, as shown below.

```{r, warning=F, message=F, fig.width=8, fig.height=4}

plot1 <- cal1020 %>%
  ggplot(aes(x=age, y=speed.mph, fill=sex, color=sex, group=sex)) + 
  geom_point(size=0.5, alpha=0.5, position=position_jitter(0.25))+
  geom_smooth(method='lm',alpha=0.5)+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  theme_minimal()+
  theme(legend.position = 'none')

plot2 <- cal1020 %>%
  ggplot(aes(x=age, y=speed.mph, fill=sex, color=sex, group=sex)) + 
  facet_grid(~sex)+
  geom_point(size=0.5, alpha=0.5, position=position_jitter(0.25))+
  geom_smooth(method='lm',alpha=0.5)+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  theme_minimal()+
  theme(legend.position = 'none')

gridExtra::grid.arrange(plot1, plot2, nrow=1, widths = c(2, 3))
```


## categorical ~ numerical

It's not very common to see a category distribution as a function of a numerical variable in experimental data, because such visualizations look like noise unless without a *lot* of data.  However, the two obvious options are variations of a stacked area plot.  I recommend the "filled" variety, unless the distribution of the explanatory variable is *really* important to show.

```{r, warning=F, message=F, fig.width=7, fig.height=3}
plot1 <- ggplot(cal1020, aes(x=round(speed.mph*2)/2, fill=sex))+
  geom_bar(position='stack')+
  scale_x_continuous('Avg speed (mph)', breaks = seq(0, 15, by=1))+
  theme_minimal()+
  theme(legend.position = 'none')
plot2 <- ggplot(cal1020, aes(x=round(speed.mph*2)/2, fill=sex))+
  geom_bar(position='fill')+
  scale_x_continuous('Avg speed (mph)', breaks = seq(0, 15, by=1))+
  ylab('proportion')+
  theme_minimal()
gridExtra::grid.arrange(plot1, plot2, nrow=1, widths = c(3,4))
```

## 2 x categorical and categorical ~ categorical

### Heatmap

To show the joint distribution of two categorical variables, the best option is a heatmap.

```{r, warning=F, message=F}
plot1 <- spsp %>% 
  mutate(stage = factor(stage, levels = c("Undergrad", "Grad", "Early Career", "Regular Member", "Retired"))) %>%
  ggplot(aes(x=stage, y=ethnicity))+
      geom_bin2d() +
  scale_fill_continuous(low="#DDDDDD", high="#000000")+
  theme_minimal()+
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90, hjust = 1, vjust=1))
```


### categorical ~ categorical

To show how distribution of one categorical variable changes as a function of another, we should use filled stacked area plots.  Note, that since these plots are "filled" along y, we get no sense of the distribution of the x variable, but in such cases we generally care much less about the distribution of x, then about the conditional distribution of y|x.

```{r, warning=F, message=F}
plot1 <- spsp %>% 
  mutate(stage = factor(stage, levels = c("Undergrad", "Grad", "Early Career", "Regular Member", "Retired"))) %>%
  ggplot(aes(x=stage, fill=ethnicity))+
      geom_bar(position='fill') +
  scale_fill_brewer(palette = "Set1")+
  theme_minimal()+
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle=90, hjust = 1, vjust=1))
```

## Extra plot notes.

### numerical ~ 2 x categorical

When plotting one numerical variable as a function of two categorical variables, the most common method is to use color to differentiate adjacent categories on the x-axis.

```{r, warning=F, message=F}

plot1 <- cal1020 %>% 
  mutate(first.1 = substr(name.first, 1, 1)) %>%
  ggplot(aes(x=sex, fill=first.1, color=first.1, y=speed.mph))+
  geom_violin(alpha=0.3,
               position="dodge")+
  # geom_jitter(width=0.25, size = 0.75, alpha=0.7,
  #              position=position_dodge())+
  stat_summary(fun.data = mean_se,
               geom="pointrange", 
               fatten = 2,
               position=position_dodge(width=0.9),
               size=1)+
  scale_y_continuous('Speed (mph)', breaks = seq(0, 15, by=1))+
  theme_minimal()+
  theme(legend.position = 'none')

```





#### Frequency vs Proportion splots

When showing how a categorical variable changes as a function of another variable, we use stacked area plots of one sort or another.  We have a choice about whether to 'stack' or 'fill' the counts.  Meaning, do we show the proportion of category y, without showing the distribution of category x (if we fill), or do we do a worse job conveying the proportions of y, but give a better sense of the distribution of x (if we stack).  Generally, the correct answer is to fill.

```{r, warning=F, message=F}
plot1 <- ggplot(spsp, aes(x=ethnicity, fill=ethnicity)) + 
  geom_bar()+
  ## the lines above form the basis of the plot, 
  ## the lines below just make it look nicer
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90, hjust=1))
plot2  <- ggplot(spsp, aes(x=ethnicity, fill=ethnicity)) + 
      geom_bar(aes(y = (..count..)/sum(..count..)))+
  ## the lines above form the basis of the plot, 
  ## the lines below just make it look nicer
  ylab('proportion')+
  scale_fill_brewer(palette = "Set1")+
  theme_bw()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle=90, hjust=1))
```


### bin width and bandwidths

The plots below will show how histograms and density plots change as a function of bin/band width.

```{r, warning=F, message=F}
plots = list()
i = 1
for(W in c(1, 5, 15)){
  plots[[i]] <- ggplot(cal1020, aes(x=time.sec/60))+
  geom_histogram(binwidth = W)+
  scale_x_continuous('Minutes', breaks=seq(0,240, by=30))+
  theme_minimal()
  i <- i+1
}

for(W in c(1, 5, 15)){
  plots[[i]] <- ggplot(cal1020, aes(x=time.sec/60))+
  geom_density(bw = W, fill='gray', alpha=0.5)+
  scale_x_continuous('Minutes', breaks=seq(0,240, by=30))+
  theme_minimal()
  i <- i+1
}

#do.call(gridExtra::grid.arrange, c(plots, ncol=3))
```